---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

The `mcycle` data regard a series of measurements of head acceleration in a simulated motorcycle accident, used to test crash helmets. The observations consist of acceleromter readings taken through time, thus there are only two variables `times`, in milliseconds after impact, and `accel`, measured in g, which can be positive or negative with 0 as a resting state.

```{r load_mcycle}
library(tidyverse)
data(mcycle, package = 'MASS')

qplot(data=mcycle, x=times, y=accel)
```

Now fit a regression model of your choosing (e.g. using `lm` `glm` or your favorite package) and plot your predictions against it.  

```{r model0, eval=FALSE}
model_start = modelfunc(?, data = mcycle)
  
library(modelr)  # install if you don't have
mcycle = mcycle %>% 
  add_predictions(model_start, var = 'prediction')

mcycle %>% 
  ggplot(aes(times, accel)) +
  geom_point() +
  geom_line(aes(y=prediction))
```

Now let's let `ggplot` give us a fit.

```{r geom_smooth}
mcycle %>% 
  ggplot(aes(times, accel)) +
  geom_point() + 
  geom_smooth(method = 'gam', formula = y ~ s(x)) # the y and x are just generic names for the axes
```

Pretty good!  Under the hood we've told ggplot to use gam from the mgcv package.  The following shows what this might look like had we done it ourselves.

```{r gam}
library(mgcv)   # part of base R so you already have it

model_gam = gam(accel ~ s(times), data = mcycle)
summary(model_gam)
```

The syntax is pretty easy, which is great!  This means you can more or less jump right in with your current R modeling knowledge.  However, know that there are many more possibilities for us in the future.

The output is confusing, but we'll understand it better later.  For now, we can see that the *smooth* term (i.e. `times`) is statistically notable, so we can feel okay about trying to understand its effects.  So let's visualize some predictions.


```{r vis_gam}
mcycle %>% 
  add_predictions(model_gam, var = 'gam_predictions') %>% 
  ggplot(aes(times, accel)) +
  geom_point() + 
  geom_line(aes(y = gam_predictions), color = "#b2001d", size = 1.2)
```


This clearly isn't a linear relationship, and it's not straightforward in general.  THAT'S OK! That's just the way nature is.  

In this case, we don't see any change until about 15 ms, at which point gs go sharply negative, followed by a rebound and eventually settling back to the resting state (though with some oscillation).  