---
title: "Issues"
output: 
  html_notebook: 
    code_folding: none
    highlight: pygments
    theme: sandstone
editor_options: 
  chunk_output_type: inline
---

## Initialization
This assumes the prior Rmd files have been run.  See the README file.

```{r misc_functions}
source('misc_functions/functions.R')
```

```{r load_packages}
library(tidyverse)
library(mgcv)
```

## Choice of Smoothing Function

There are a lot of choices, but the default is generally good enough.  There are specific cases for categorical factors, spatial random effects, etc., but you only use those when appropriate.

```{r smooth_func}
# cubic spline, gaussian process, and b-spline; see ?smooth.terms
mod_gam_s <- gam(Overall ~ s(Income, bs = 'cs') + s(Edu, bs = 'gp') + s(Health, bs = 'ps'), data=pisa)
summary(mod_gam_s)
```

A very common setting would be to allow smooth terms to vary over groups.

```{r by}
pisa_math_sci = readr::read_csv('data/pisa_2000_thru_2015_math_science.csv')
pisa_math_sci = pisa_math_sci %>% 
  filter(!grepl(Jurisdiction, pattern = 'Average')) %>% 
  spread(key = test, value = Average) %>% 
  mutate(Year_fac = factor(Year))

mod_math_sci = gam(Science ~ s(Math, by = Year_fac), data = pisa_math_sci)

plot(mod_math_sci)
# visibly::plot_gam_by(mod_math_sci, main_var = Math, by_var = Year_fac)


# another example
library(lme4)
mod_reaction = gam(Reaction ~ s(Days, by = Subject), data = sleepstudy)
plot(mod_reaction)
# visibly::plot_gam_by(mod_reaction, main_var = Days, by_var = Subject)
```


## Diagnostics

The `gam.check` function mimics `lm` as far as the plots go.  Additionally, it provides some information as to whether you have allowed the term to be 'wiggly enough'.

```{r diagnostics}
gam.check(mod_gam2)  # mod_gam2 is from application.Rmd
```

## Prediction

We can get predictions via the standard `predict` function, but packages like `modelr` and `broom` can make it easy to add to current data for plotting.

```{r predict_modelr}
library(modelr)

pisa %>% 
  add_predictions(mod_gam1)
```

```{r predict_broom}
library(broom)

augment(mod_gam1)
```




## Model Comparison

We can compare models with AIC or an anova test where appropriate.  In the following, the `ti` allows for a more standard 'main effects plus interaction' approach.

```{r model_comparison}
mod_A = gam(Overall ~ s(Income, bs="cr", k=5) + s(Edu, bs="cr", k=5), data=pisa)
mod_B = gam(Overall ~ ti(Income, bs="cr", k=5) + ti(Edu, bs="cr", k=5) + ti(Income, Edu, bs='cr'), data=pisa)

AIC(mod_A, mod_B)
anova(mod_A, mod_B, test="Chisq")
```

