---
title: "Issues"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r misc_functions}
source('misc_functions/functions.R')
```

```{r load_packages}
library(tidyverse)
library(plotly)
library(modelr)
library(mgcv)
```

## Choice of Smoothing Function

There are a lot of choices, but the default is generally good enough.  There are specific cases for categorical factors, spatial random effects, etc., but you only use those when appropriate.

```{r smooth_func}
# cubic spline, gaussian process, and b-spline; see ?smooth.terms
mod_gam2 <- gam(Overall ~ s(Income, bs = 'cs') + s(Edu, bs = 'gp') + s(Health, bs = 'ps'), data=pisa)
summary(mod_gam2)
```

```{r by}
mod_gam2 <- gam(Overall ~ s(Income, bs = 'cs') + s(Edu, bs = 'gp') + s(Health, bs = 'ps'), data=pisa)
summary(mod_gam2)


```


## Diagnostics

The `gam.check` function mimics `lm` as far as the plots go.  Additionally, it provides some information as to whether you have allowed the term to be 'wiggly enough'.

```{r diagnostics}
gam.check(mod_gam2)
```

